---
title: "BCB_Project"
author: "Chiteri and Tanner and Juan"
date: "4/4/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Dependencies

library("tidyverse")
library("factoextra")
library("FactoMineR")
library("readxl")


library("ggfortify")
library("FactoMineR")
library("princurve")

```


#Reading data

```{r}
# Counts
rna <- read_csv('./Data/GSE121039_AR_scRNAseq_transcript_counts.csv')
rna <- rename(rna, gene = X1) #rename first column

# Quality control: PASS/FAIL

qc <- readxl::read_excel('./Data/aav6428_Nelms_Table-S1.xlsx', sheet = 'Table S1')

names(qc)[12] <- 'QC' 
qc <- qc[qc$QC == "PASS", "Cell Number"]

# Deferentially expressed genes during specific stages
deg <- readxl::read_excel('./Data/aav6428_Nelms_Table-S2.xlsx')[,1:2]
names(deg) <- c('gene', 'phase')
deg <- deg %>% filter(phase != "G1") %>% select(gene)

head(rna)
head(qc)
```

```{r}
#get into long format
rna1 <- pivot_longer(rna, cols = starts_with('Cell'))
```
#rename the columns
#select the columns


```{r}
#separate 'name' into 4 columns, 

rna3 <- separate(rna1, name, into = paste0("y",1:4), sep = " ") %>% 
  select(!c(y1, y3)) %>% rename(c('cell' = y2, 'rep'= y4, 'count' = value))

```


```{r}
#Filter cells based on QC
rna4 <- filter(rna3, cell %in% qc$`Cell Number`)
names(rna4)[1] <- 'gene'
head(rna4)

paste0("Total number of cells after Quality Control: ", length(unique(as.numeric(rna4$cell))))
```

``` {r}

##dataset to use for e.g. Fig. 2A, Fig S4). We need 128 cells
#pooling cells using sum
rna5 <- rna4 %>%  group_by(gene,cell,) %>% summarise(n=n(), sum = sum(count))
paste("There are ", length(unique(as.numeric(rna5$cell))), " unique cells in rna5")

```

``` {r}
rna6 <- filter(rna5, n > 1)

paste0("There are ", length(unique(as.numeric(rna6$cell))), " cells with a technical replicate in rna6.")
```


``` {r}
#Join two columns cell and rep and go wider

rna7 <- rna4 %>%  
  unite("cell",c(cell, rep), sep = "_") %>% 
  pivot_wider(names_from = cell, values_from = count, names_prefix = 'cell')

dim(rna7)
```


``` {r}
#remove gene title
rna9 <- rna7 %>% remove_rownames %>% column_to_rownames(var = "gene")

dim(rna9)

rna7[1:5, 1:5]
rna9[1:5, 1:4]

```

```{r}
# transpose data frame so cells are rows and genes are columns
rna9_t <- as.data.frame(t(as.matrix(rna9)))
```

```{r}
# next two chunks for getting rid of 'unreplicated' cells 
cells <- as.data.frame(row.names(rna9_t))
names(cells) <- 'key'

cells %>% 
  separate(key, c('cell', 'rep'), remove = F) %>% 
  group_by(cell) %>% filter(n() == 2) %>% 
  ungroup() %>% select(key) -> cells_after_qc
```


```{r}
# filtering out rna9 after quality control
rna9_t %>% rownames_to_column() %>% 
  filter(rowname %in% cells_after_qc$key) %>% 
  column_to_rownames() -> rna9_f
```


```{r}
# Re arranging rna9_aqc moving cells as columns 
rna9_ft <- as.data.frame(t(as.matrix(rna9_f)))
rna9_ft %>% rownames_to_column("gene") -> rna9_ft 
```

```{r}
# Filtering out genes with less than 100 counts across cells 
rna9_ft %>%   
  arrange(gene) %>% rowwise() %>%  
  group_by(gene) %>% dplyr::summarise(m = sum(c_across(2:length(rna9_ft)-1))) %>%  
  filter(m > 100) %>% select(gene) -> exp_genes

rna_ftf <- as_tibble(rna9_ft[rna9_ft$gene %in% exp_genes$gene,])
dim(rna_ftf)
```

#Filtering out genes differentially expressed in specific stages
```{r}
rna_ftfc <- rna_ftf %>% filter(!gene %in% deg$gene)
```

Normalization
```{r}

# Step 1: divided each individual count by the total counts per cell, and multiply by 1000000 

rna_norm <- as_tibble(rna_ftfc$gene) # tibble ngenes x 1 for starting fillling with normalized data

for (i in 2:length(rna_ftfc)) {
  a <- (rna_ftfc[,i]/sum(rna_ftfc[,i]))*1000000
  rna_norm <- cbind(rna_norm, a)
  a <- NULL
}

rna_norm <- column_to_rownames(rna_norm, 'value') # for transposing

# Step 2: log transformation after add 11

rna_log <- log10(rna_norm[,1:length(rna_norm)] + 11 )

rna_log[1:5, 1:5]

rna_final <- as.data.frame(t(as.matrix(rna_norm)))
rna_final[1:5, 1:5]

```


```{r}
PCA = prcomp(rna_final, center=TRUE, scale.=FALSE)
summary(PCA)

PCA$x %>% 
  as.data.frame %>%
  ggplot(aes(x = PC1, y = PC2, size=100)) + 
  geom_point()+theme(legend.position = "None")

ggsave('pca2.jpeg')
```

#Figures assignment
Fig 1c
Fig2 a b c d
Fig3 a b d
Fig4 a b c d e
Fig5 c e

##Sean 3 a b d
```{r}

```


##Tanner 2 a c d




```{r}
#Dependencies
library("princurve")
library(data.table)
library(pheatmap)

```


```{r}

####Pseudotime

##Take Normalized genes and the filter with 2000 genes with highest variance and use for pseudotime. From fig. S3

 Anther_size = read.table("./Data/Updated_S1.txt", header= T)

#Transform data to merge files to use only 2000 genes
rna_final_Tanner= as.data.frame(t(rna_final)) 
s3= read.csv("./Data/aav6428_Nelms_Table-S3.csv", header=T)

#Add a column label to rownames
setDT(rna_final_Tanner, keep.rownames = "gene")[]

#Merge Files
Variance2000transcripts=inner_join(rna_final_Tanner,s3 )### MAke a combined table of all fig s3, order by variance and then slice 2000

y= Variance2000transcripts %>% arrange(desc("Gene score (% variance explained by pseudotime)")) %>% 
  select(1:258) %>% 
  top_n(2000)

 
#Remove gene id column
y$gene=NULL

#Trying the PCA with prcomp

PCA2= prcomp(y, center=TRUE, scale.=FALSE)
summary(PCA)
PCA2$x %>% 
  as.data.frame %>%
  ggplot(aes(x = PC1, y = PC2)) + 
  geom_point()

plot(PCA2)
names(PCA2)

#PCA 10 as a matrix while selecting the X values from the PCA
PCAX= as.matrix(PCA2$x [,1:10])
PCAX

I=principal_curve(PCAX)
names(I)

#Principal Curve
plot(I$s, xlim=c(-20,10))
I$s

#Overlay principle curve for PCA in fig.1. Makes Fig. 2A
df1=as.data.frame(PCA$x)
df2=as.data.frame(I$s)

df1 %>% 
  ggplot(aes(x =PC1, y = PC2)) + 
  geom_point() +
  geom_smooth(data = df2, aes(x = PC1, y = PC2)) +
  theme(legend.position = "none")+xlim(-20,30)


ggsave('Pseudotime_PCA_overlay.jpeg')
```


```{r}
#### Heatmap  Fig. 2
#Sort by Anther size

ANTHER1=read.csv("./Data/with_anther_size", header=T)


ANTHER2=ANTHER1[order(ANTHER1$anther_size_mm),]
ANTHER2=as.tibble(ANTHER2)

#Remove cells that don't pass QC Result=144 cells
ANTHER3 <- filter(ANTHER2, cell %in% qc$`Cell Number`)
ANTHER4=t(ANTHER3)

#Remove rows X, cell, rep,anther_size_mm
ANTHER5=ANTHER4[-c(1,2,3,4),]
ANTHER6=as.data.frame(ANTHER5)

#Add row names as a column for sorting

setDT(ANTHER6, keep.rownames = "gene")[]

#Anther length with 3040 genes
Anther_length_w_genes=inner_join(ANTHER6,s3) # 3040 genes and 144 cells



#Remove gene names 
Anther_length_w_genes$gene=NULL

#Make heatmap ordered by anther length
Anther_length_w_genes1=(t(Anther_length_w_genes))


vst_cor<- as.matrix(cor(Anther_length_w_genes1, method="spearman"))
vst_cor2<- as.matrix(cor(Anther_length_w_genes, method="spearman"))


pheatmap(vst_cor,border_color="white",labels_col =NA,labels_row = "3040 genes", treeheight_row = 0, treeheight_col = 0, cluster_rows = 2, cluster_cols = 2)
pheatmap(vst_cor2,border_color="white",labels_col =NA, labels_row = NA, treeheight_row = 0, treeheight_col = 0, cluster_rows = 2, cluster_cols = 2)


```

##Yuru 5a,5c
```{r}
### Figure 5A

### extract the expresses genes information in the table_S2
Table_S2 <- read_excel("./Data/aav6428_Nelms_Table-S2.xlsx")
#str(Table_S2)
colnames(Table_S2) <- c("gene", "Cell_cycle_phase",  "Meiotic_expression_level", "Gene_discription")
Table_S2$Cell_cycle_phase <- as.factor(Table_S2$Cell_cycle_phase)
dim(Table_S2)
MEG <- Table_S2$gene ## Expresses genes in meiotic cycle 
length(MEG)


rna_ftfc1 <- rna_ftf
rna_norm1<- as.tibble(rna_ftf$gene)
for (i in 2:length(rna_ftf)) {
  a <- (rna_ftf[,i]/sum(rna_ftf[,i]))*1000000
  rna_norm1 <- cbind(rna_norm1, a)
  a <- NULL
}
rna_norm1 <- column_to_rownames(rna_norm1, 'value')
rna_norm1 <- log10(rna_norm1[,1:length(rna_norm1)] + 11 )
rna_norm_456 <- rna_norm1 %>% rownames_to_column(var = "Gene") %>% filter(Gene %in% MEG) #%>% select(!Gene)
dim(rna_norm_456)

### PCA of 448 genes after filtering MEG out from rna_norm1
PCA_figure5a = prcomp(rna_norm_456[,-1], center=TRUE, scale.=TRUE)

### PCA plotting
PCA_figure5a$x %>% 
  as.data.frame %>%mutate(phase=Table_S2.1$Cell_cycle_phase) %>% 
  ggplot(aes(x = PC1, y = PC2)) + 
  geom_point(aes(col=phase)) +
  labs(title="Principal component plot of cell cycle–regulated genes", subtitle="Figure 5A")
ggsave("./Figures/Figure5_A.png")

### Figure 5C
Table_S1 <- read_excel("./Data/aav6428_Nelms_Table-S1.xlsx",sheet = "Table S1")
dim(Table_S1)
colnames(Table_S1)
Table_S1.1 <- Table_S1[,c(1, 3, 13, 14)]
colnames(Table_S1.1) <- c("cell", "Anther_length", "Expression_stage", "Cell_cycle_phase")
Table_S1.1$cell <- as.factor(Table_S1.1$cell)
Table_S1.1$Cell_cycle_phase <- as.factor(Table_S1.1$Cell_cycle_phase)

cbp1 <- c("#999999", "#56B4E9", "#C3D7A4","#009E73","#293352", "#0072B2")
Table_S1.1 %>% filter(!Cell_cycle_phase == "N.D.") %>% droplevels() %>% 
  mutate(anther_length = cut(Anther_length, breaks= c(0.2, 0.4, 0.6, 0.85, 1.1, 1.25, 1.4, 1.6))) %>%
  ggplot()+
  geom_bar(aes(x=anther_length, fill=Cell_cycle_phase), position = "fill", width=0.5)+
  scale_fill_manual(values = cbp1) +
  labs(title="Estimated proportion of cells assigned to each cell-cycle", subtitle="Figure 5C")
ggsave("./Figures/Figure5_C.png")


```

##Chiteri 1 c 2 b 5e
```{r}
# Quality control: PASS/FAIL
table1 <- readxl::read_excel('./Data/aav6428_Nelms_Table-S1.xlsx', sheet = 'Table S1')

#rename the columns..replace spaces with underscore and lower case
table1.1 <- table1 %>%  
  select_all(~gsub("\\s", "_", .)) %>% 
  select_all(tolower)

#rename anthersize column name
colnames(table1.1)[3] <- 'anther_size_mm'
colnames(table1.1)[1] <- 'cell'

#filter only cells that passed

table1.2 <- table1.1 %>%  filter(quality_control=="PASS") %>%  select(cell, anther_size_mm)

#Join o wider

rna4_wide <- rna4 %>% pivot_wider(names_from = gene, values_from = count)

rna4_anther <- merge(rna4_wide, table1.2, by='cell') %>% select(cell, rep, anther_size_mm, everything()) %>% arrange(as.numeric(cell))

rna4_replicated <- rna4_anther %>% group_by(cell) %>% filter(n()==2) %>%  ungroup() 


#make anther_size_mm for grouping purposes
rna4_replicated$anther_size_mm <- as.character(rna4_replicated$anther_size_mm)

#filter whose sum across samples is > 100
rna4_filtered <- rna4_replicated[4:ncol(rna4_replicated)][, apply(rna4_replicated[4:ncol(rna4_replicated)],2,sum) > 100]


#we get 256 obs, 12912 genes, 10 shy of 12902 from the paper

#introduce the cell, rep, anther_size
rna4_x <- cbind(rna4_replicated[1:3], rna4_filtered )

#filter the 375 
# Deferentially expressed genes during specific stages
rna4_x2 <- rna4_x[, !names(rna4_x) %in% deg$gene] 
#we get 12544 against 12527 in the paper. We have 17 more??


#make anther_size_mm for grouping purposes
rna4_x2$anther_size_mm <- as.numeric(rna4_x2$anther_size_mm)

#Uses row sums
rna4_x3 <- apply(rna4_x2[4:ncol(rna4_x2)], 1,function(i) log10(((i/sum(i))*1000000)+11 ))

rna4_log3 <- as.data.frame(t(rna4_x3))

#combine
rna4_log3 <- cbind(rna4_x2[,1:3], rna4_log3)

PCA_256 = prcomp(rna4_log3[, -c(1:3)], center=TRUE, scale.=FALSE)
#summary(PCA_256)

#size=8 Will make the dots better. Place before colour


ggfortify::autoplot(PCA_256, data=rna4_x2, size=5, colour= 'anther_size_mm') + geom_line(aes(group = cell))+ scale_color_gradient2(low = "gray",high = "black", space = "Lab" )+theme_bw()

ggsave('pca2.png')


autoplot(PCA_256, data=rna4_x2, size=2, colour= 'anther_size_mm') + geom_line(aes(group = cell))+ scale_color_gradient2(low = "gray",high = "black", space = "Lab" )+theme_bw()



ggsave('./Figures/pca2.png')

```
##Attempt on clustering
```{r}
#filter whose sum across samples is > 100
rna4_272 <- rna4_anther[4:ncol(rna4_anther)][, apply(rna4_anther[4:ncol(rna4_anther)],2,sum) > 100]

#we get 272 obs, 13187 genes, 285 shy of 12902 from the paper

#introduce the cell, rep, anther_size
rna4_272x <- cbind(rna4_anther[1:3], rna4_272 )

#filter the 375 
# Deferentially expressed genes during specific stages
rna4_272xx <- rna4_272x[, !names(rna4_272x) %in% deg$gene] 
#we get 12817 against 12527 in the paper. We have 290 more??

# extract data that is for clustering on Deferentially expressed genes during specific stages
cell_cycle <- readxl::read_excel('./Data/aav6428_Nelms_Table-S2.xlsx')[,1:2]
names(cell_cycle) <- c('gene', 'phase')



# Grab Deferentially expressed genes during specific stages
rna4_c1 <- rna4_272x[, names(rna4_272x) %in% cell_cycle$gene] 
#Grabs the 453 variables we need
#3 genes not found in here

#add cell, rep, anther length
rna4_c2 <- cbind(rna4_272xx[, 1:3], rna4_c1)

##normalize
#divide cell with each row sum
rna4_c3 <- apply(rna4_c2[4:ncol(rna4_c2)], 1,function(i) log10(((i/sum(i))*1000000)+11 ))



rna4_c4 <- as.data.frame(t(rna4_c3))

#combine
rna4_c5 <- cbind(rna4_272xx[,1:3], rna4_c4)
#write.csv(rna4_274b, file="with_anther_size")

PCA_c6 = prcomp(rna4_c5[, -c(1:3)], center=TRUE, scale.=TRUE)
#summary(PCA_c6)

#cell-cycle gene by anther_size
autoplot(PCA_c6, data= rna4_c5, colour = 'anther_size_mm')+ 
  theme_bw() 

```

```{r}
##kmeans with cycle

#kmeans plot
set.seed(20)
autoplot(kmeans(rna4_c5[, -c(1:3)], 6), data = rna4_c5)
ggsave('./Figures/cell_cycle_cluster.png')

```

```{r}
#fitting principle curve
#getting the matrix to dataframe
#get first 10PCA
c <- as.data.frame(PCA_256$rotation[, 1:10])
c1 <- rownames_to_column(c, var='gene')

pc1 <-c1 %>% head(arrange(desc(PC1)), n=2000) %>% select(gene)
pc2 <-c1 %>% head(arrange(desc(PC2)), n=2000) %>% select(gene)
pc3 <-c1 %>% head(arrange(desc(PC3)), n=2000) %>% select(gene)
pc4 <-c1 %>% head(arrange(desc(PC4)), n=2000) %>% select(gene)
pc5 <-c1 %>% head(arrange(desc(PC5)), n=2000) %>% select(gene)
pc6 <-c1 %>% head(arrange(desc(PC6)), n=2000) %>% select(gene)
pc7 <-c1 %>% head(arrange(desc(PC7)), n=2000) %>% select(gene)
pc8 <-c1 %>% head(arrange(desc(PC8)), n=2000) %>% select(gene)
pc9 <-c1 %>% head(arrange(desc(PC9)), n=2000) %>% select(gene)
pc10 <-c1 %>% head(arrange(desc(PC10)), n=2000) %>% select(gene)

#create a dataframe with the top 2000 most vairable
top2000 <- data.frame(pc1, pc2, pc3,pc4,pc5,pc6,pc7,pc8,pc9,pc10)

top2k <- top2000$gene

#filter original transformed data
top_2000a <- rna4_x2[, names(rna4_x2) %in% top2k] 

#top_2000a <- rna4_x2 %>% select(cell, rep, anther_size_mm, names(rna4_x2) %in% top2k)

#Uses row sums
top_2000b <- apply(top_2000a[1:ncol(top_2000a)], 1,function(i) log10(((i/sum(i))*1000000)+11 ))

top_2000c <- as.data.frame(t(top_2000b))

#combine,  find right way to get cell, rep, anther
#top_2000d <- cbind(rna4_x2[,1:3], top_2000c)

top_pca = prcomp(top_2000c[, 1:ncol(top_2000c)], center=TRUE, scale.=TRUE)

autoplot(top_pca)

##principle curve
pc <- principal_curve(as.matrix(top_2000c[, 1:ncol(top_2000c)]))
plot(pc)

pcx <- as.matrix(PCA_256$rotation[, 1:10])
I=principal_curve(pcx)
plot(I)

```



##Juan 4b d e
# Figure 4b

Getting data for figures 4 b

```{r}
# Required data frames. From "Table.S4" we need gene names for Clusters 5 and 6  

Table.S4 <- readxl::read_excel('./Data/aav6428_Nelms_Table-S4.xlsx')
names(Table.S4) <- c('Gene', 'Category', 'Gene_name', 'Cluster', 'Meiocytes_Exp') 

cluster_5 <- Table.S4 %>% filter(Cluster == 5) %>% select(Gene)
cluster_6 <- Table.S4 %>% filter(Cluster == 6) %>% select(Gene)

# Load Table.S1, from we will anther length

Table.S1 <- readxl::read_excel('./Data/aav6428_Nelms_Table-S1.xlsx', sheet = "Table S1")

Table.S1 %>% 
  filter(!`Cell cycle phase` == "N.D.") %>% droplevels() %>% 
  select(`Cell Number`, `Plant Number`, `Cell cycle phase`, `Anther size (mm)`) %>% 
  arrange(match(`Cell cycle phase`, c("G1", "G1/S", "S", "G2/M", "M", "mei")))


```
Dataframe for Figure 4b
Figure 4 b plots Transcripts per million vs pseudotime. Given Pseudotime was not obtained,
we used anther length as a proxy of the developmental stage.

To re create this figure, first we need to pick up significant genes in Clusters 5 and 6 
from the normalized counts, before log transformation.

```{r}
rna_norm %>% 
  rownames_to_column("gene") %>% 
  filter(gene %in% c(cluster_5$Gene, cluster_6$Gene)) %>% 
  column_to_rownames("gene") %>% t() %>% 
  as.data.frame() %>% 
  rownames_to_column('Cell') %>% 
  separate(Cell, into = c('Cell Number', 'Rep'), sep = "_") -> fig4
```

After filtering genes, column `Cell Number` is modified for allowing match its name to 
Table.S1, from where we get `Anther Length (mm)`

```{r}
fig4$`Cell Number` <- parse_number(fig4$`Cell Number`) 

fig4 %>%  
  left_join(., Table.S1[,c(1, 3)], by = "Cell Number") -> fig4t

```

Plotting Afd1/Rec8 and Zyp1

```{r}
library(patchwork) # Patchwork allows a nicer syntax for combining multiple ggplots

# Define limits in anther size for Developmental stage, stored as `$Stage`
fig4t$Stage <- cut(fig4t$`Anther size (mm)`, c(0, 0.75, 1.1, 1.3, 1.4, Inf), 
    labels = c('Mitotic expansion',
               "AR/PMC transition",
               'PMC/Pr1',
               "Pr1/PR2 transition",
               'End of Early meoitic prophase'))

# Two plots combined afterwards. "Afd1/Rec8" and "Zip1";
# aes() includes three terms: Anther size (our x, continuous), 
#                             Gene name (it is actually a column of counts, continuous)
#                             Stage (categorical) 
# geom_point plots counts vs anther size
# geom_smooth plots the overlaid smoothed curve
# references added and edited by labs() and theme() options

fig4.b.1 <- fig4t %>% 
  ggplot(aes(x = `Anther size (mm)`, y = Zm00001d039133, color = Stage)) + 
  geom_point() +
  geom_smooth(color = 'red', se = F) +
  labs(title = "Afd1/Rec8", caption = "Cluster 5 \nGradual Upregulation") + xlab(NULL) + ylab("TPM") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(size = 8, face = "italic"))

fig4.b.2 <- fig4t %>% 
  ggplot(aes(x = `Anther size (mm)`, y = Zm00001d025575, color = Stage)) + 
  geom_point() +
  geom_smooth(color = 'red', se = F) +
  labs(title = "Zip1", caption = "Cluster 6 \nLate Upregulation") + xlab('Anther length (mm)') + ylab("TPM") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(size = 8, face = "italic"))

# The following code calls the patchwork's grammar for creating a merged plot and combining legends

fig4b <- (fig4.b.1 / fig4.b.2) + plot_layout(guides = 'collect')

ggsave('./Figures/Figure4_b.png', plot = fig4b)

```
# Figure 4.d

This figure was generated after running manual analysis usign by  the webtool AgriGO V2.0.
Gene names from each cluster where taken from Table.S4, then Singular Enrichment Analysis was performed by cluster, using as reference "Maize AGPv4 (Maize-GAMER)". Genes with GO terms associated to the only word
"translation" where selected for this analysis (in genes_clust.txt).

```{r}
# Call genes_clust for getting genes present in each gene cluster by unsupervised clustering

clust <- read_delim(file = './Data/genes_clust.txt', delim = '\t', col_names = T)

# Create the dataframe for figure 4d: Counts of occurrences of differentially expressed genes in each cluster

clust %>% 
  group_by(Cluster, Name) %>% 
  summarise(n = n()) %>%
  group_by(Cluster) %>%
  summarise(n = n()) %>%
  add_column(total = c(560, 507, 275, 
                       108, 412, 129)) %>%
  mutate(percent = n/total*100) -> fig4d

fig4d
```

Bar plot for gene counts in each cluster 
```{r}
f4d <- ggplot() + 
  geom_col(data = fig4d, aes(x= Cluster, y = percent, fill = 'pink')) +
  geom_hline(yintercept = (sum(fig4d$n)/3046*100), linetype = 'dashed') +
  scale_x_continuous(breaks = 1:6) +
  geom_text() +
  annotate("text", label = paste0("Expected translation GO temrms \nif distributed at random: ", 
                                  round(sum(fig4d$n)/3046*100, 2), " %"),
           x = 5, y = 13) +
  labs(title = "Figure 4.D", 
       subtitle = "Genes involved in translation GO term for each cluster", 
       y = "% translation genes") +
  theme_bw() +
  theme(legend.position = 'none', 
        plot.subtitle = element_text(face = "italic"))

f4d  
ggsave('./Figures/Figure4_d.png', plot = f4d)
```




# Additional figures for 2D

On the right hand-side of the heatmap.
The starting file has counts of rna before log transformation.
Given that genes reported in the paper turned put to be filtered out due to low expression,
we took the data file even before filtering those genes, re-normalized it, and genes of interest
were filtered

```{r}
fig2 <- as.tibble(rna9_ft) # call rna counts before filtering out low expressed genes

# TPM normalization as describe in the supplementary methods

fig2df <- as.tibble(fig2$gene)
for (i in 2:length(fig2)) {
  a <- (fig2[,i]/sum(fig2[,i]))*1000000
  fig2df <- cbind(fig2df, a)
  a <- NULL
}
```

Steps in the next chunk:
1. Filtering counts file for getting genes Rps4a, Unk1, Unk2, Rmf, Trps8 into a new datafile. 
Rpl38e Identifier not founded
2. Reshape datafile and Column `Cell Number` modified to match to Table.S1
3. Merge filtered df to anther length (mm) from Table.S1

``` {r}
# 1. filtering
fig2df <- fig2df %>%  
  filter(value %in% c('Zm00001d027037', 'Zm00001d039101', 'Zm00001d009343', 'Zm00001d050069', 'Zm00001d013377')) 

# 2. Reshaping datafile
fig2df %>% 
  column_to_rownames("value") %>% t() %>% 
  as.data.frame() %>% 
  rownames_to_column('Cell') %>% 
  separate(Cell, into = c('Cell Number', 'Rep'), sep = "_") -> fig2ddf

# 2. Modify column `Cell Number` 
fig2ddf$`Cell Number` <- parse_number(fig2ddf$`Cell Number`) 

# 3. Merge anther length to datafile 
fig2ddf %>%  
  left_join(., Table.S1[,c(1, 3)], by = "Cell Number") -> fig2ddf_sum
```


Plotting the gene expression vs anther size
```{r}
# Define breaks for different stages for a given anther size
fig2ddf_sum$Stage <- cut(fig2ddf_sum$`Anther size (mm)`, c(0, 0.75, 1.1, 1.3, 1.4, Inf), 
    labels = c('Mitotic expansion',
               "AR/PMC transition",
               'PMC/Pr1',
               "Pr1/PR2 transition",
               'End of Early meoitic prophase'))

# Four plots combined afterwards. "Rpl38e" and "Rps4a" were discarded given low counts or not counted;
# aes() includes three terms: Anther size (our x, continuous), 
#                             Gene name (it is actually a column of counts, continuous)
#                             Stage (categorical) 
# geom_point plots counts vs anther size
# geom_smooth plots the overlaid smoothed curve
# references added and edited by labs() and theme() options

Unk1 <- 
  fig2ddf_sum %>%
  ggplot(aes(x = `Anther size (mm)`, y = Zm00001d027037, color = Stage)) + 
  geom_point() +
  geom_smooth(color = 'red', se = F) +
  labs(title = "Zip1") + xlab(NULL) + ylab("TPM") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0))

  
Rmf <- fig2ddf_sum %>% 
  ggplot(aes(x = `Anther size (mm)`, y = Zm00001d039101, color = Stage)) + 
  geom_point() +
  geom_smooth(color = 'red', se = F) +
  labs(title = "Rmf") + xlab(NULL) + ylab("TPM") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0))

Rps4a <- fig2ddf_sum %>% 
  ggplot(aes(x = `Anther size (mm)`, y = Zm00001d009343, color = Stage)) + 
  geom_point() +
  geom_smooth(color = 'red', se = F) +
  labs(title = "Rps4a") + xlab(NULL) + ylab("TPM") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0))

Trps8 <- 
  fig2ddf_sum %>%
  ggplot(aes(x = `Anther size (mm)`, y = Zm00001d050069, color = Stage)) + 
  geom_point() +
  geom_smooth(color = 'red', se = F) +
  labs(title = "Trps8") + xlab(NULL) + ylab("TPM") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0))

Unk2 <- 
  fig2ddf_sum %>%
  ggplot(aes(x = `Anther size (mm)`, y = Zm00001d013377, color = Stage)) + 
  geom_point() +
  geom_smooth(color = 'red', se = F) +
  labs(title = "Unk2") + xlab('Anther length (mm)') + ylab("TPM") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0))

# The following code calls the patchwork's grammar for creating a merged plot and combining legends

fig2dextra <- (Unk1 / Rmf / Trps8 / Unk2) + plot_layout(guides = 'collect')

ggsave('./Figures/Figure2_d.png', plot = fig2dextra)
```

